Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-02 joint_inf.tex
Minor changes and suggestions; draft response.
fix deprecated $$; tex kinda sux :(

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{showlabels}
\usepackage{palatino}
\usepackage{hyphenat}
\usepackage{thm-restate}
\usepackage{lineno}
\linenumbers
\usepackage{natbib}
\usepackage{setspace}
\onehalfspacing
\usepackage{authblk}
\usepackage{fancyhdr}

@@

\fancyhead[R]{ submission intended as Article in Discoveries section of \emph{MBE} }

@@

\author[1]{David A. Shaw}
\author[1]{Frederick A. Matsen IV\thanks{Corresponding author, email: \url{matsen@fredhutch.org}}}
\affil[1]{Computational Biology Program, Fred Hutchinson Cancer Research Center\\ Seattle, WA, USA}
\date{}

@@

\maketitle
\thispagestyle{fancy}

@@

Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch fidelities $t^*$.
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
$$
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
$$
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating Farris topology ($\tau_1$).
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
$$
$$
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
$$
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

@@

We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the notion of fidelities to analyze what happens as changes occur along each branch as opposed to the standard branch length formulation.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we estimate $x$ and $y$ correctly, then, for
$$
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
$$
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
$$
to find a region where the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
$$
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
\end{restatable}
This theorem allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, if there is no change along a branch with fidelity $\theta$, the likelihood will more likely include a term of $(1+\theta)$ instead of $(1-\theta)$ for this event, resulting in positive bias of branch fidelities.
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

@@

The general case is more challenging to analyze and so we obtain weaker bounds.
Here, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
We use this theorem to show incorrect branch parameter estimates as follows.
If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{showlabels}
\usepackage{palatino}
\usepackage{hyphenat}
\usepackage{thm-restate}
\usepackage{natbib}
\usepackage{setspace}
\onehalfspacing
\usepackage{authblk}

% Line numbers and some weirdness
\usepackage{lineno}
\linenumbers
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
  \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
  \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
  \renewenvironment{#1}%
     {\linenomath\csname old#1\endcsname}%
     {\csname oldend#1\endcsname\endlinenomath}}% 
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
  \patchAmsMathEnvironmentForLineno{#1}%
  \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

% commented for arXiv submission
%\usepackage{fancyhdr}
% \fancyhead[R]{ submission intended as Article in Discoveries section of \emph{MBE} }

@@

@@

\author[1]{David A. Shaw}
\author[1]{Frederick A. Matsen IV\thanks{Corresponding author. Email: \url{matsen@fredhutch.org}}}
\affil[1]{Computational Biology Program, Fred Hutchinson Cancer Research Center\\ Seattle, WA, USA}
\date{}

@@

\maketitle
% commented for arXiv submission
%\thispagestyle{fancy}

@@

Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch fidelities $t^*$.
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
\]
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating Farris topology ($\tau_1$).
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
\[
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
\]
\[
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
\]
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

@@

%EM Line numbering has a huge gap here. Could you look around and see if you see anything about this on the web?
We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the notion of fidelities to analyze what happens as changes occur along each branch as opposed to the standard branch length formulation.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we estimate $x$ and $y$ correctly, then, for
\[
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
\]
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
\[
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
\]
to find a region where the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
\[
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
\]
\[
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
\]
The maximum likelihood value $\hat{w}$ is equal to 1 if
\[
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
\]
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
\end{restatable}
This theorem allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, if there is no change along a branch with fidelity $\theta$, the likelihood will more likely include a term of $(1+\theta)$ instead of $(1-\theta)$ for this event, resulting in positive bias of branch fidelities.
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

@@

The general case is more challenging to analyze and so we obtain weaker bounds.
Here, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
\[
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
\]
\[
\gamma := \gamma(x^*, y^*),
\]
bounds
\[
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
\]
\[
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
\]
and
\[
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
\]
the maximum likelihood value $\hat{w}=1$ when
\[
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
\]
\end{restatable}
We use this theorem to show incorrect branch parameter estimates as follows.
If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
\[
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
\]
and
\[
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
\]
Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.
